Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-20 paper/main.tex
rename video to episode

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 scenes, and treated each window as a single ``document" for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used hand-annotated descriptions of each moment of video to fit a topic model.  Three example video frames and their associated descriptions are displayed (top two rows).  Participants later recalled the video (in the third row, we show example recalls of the same three scenes from participant 17).  We used the topic model (fit to the annotations) to estimate topic vectors for each moment of video and each sentence the participants recalled.  Example topic vectors are displayed in the bottom row (blue: video annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively).  We also show the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% add an x-label under the "

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number (typically one or two) of topics tended to be ``active" in any given timepoint (see Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events" into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each video event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 video events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants, see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}
% I'm all for citing software packages, but where what's the distinction between packages we do cite (seaborn, brainiak, sklearn, hypertools, quail, umap, etc.) and don't cite (wordcloud, nilearn, fastdtw, etc.)?
% - Paxton
% I also flagged the seaborn reference for removal here. Def not critical to understanding the figure - Andy

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the video, to obtain similar estimates for each participant's recall, we treated each overlapping  window of (up to 10) sentences from their transcript as a ``document," and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participants' recalls to events from the original video, despite different participants using widely varying language to describe the events, and that those descriptions often diverged in content and quality from the video annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the video (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall trajectory (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we leveraged an HMM-based optimization procedure (see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail---i.e., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the video correlation matrix, there were substantial off-diagonal correlations.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events via HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented" events from the original video, and a (potentially differently ordered) sequence of ``recalled" events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observed patterns comparable to classic effects from list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). In contrast, we did not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather greater memory for specific events distributed approximately evenly throughout the video.

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.~Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean clustering score: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean clustering score: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.
% NOTE:
% Semantic and temporal clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.
% - Paxton

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the video. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the video.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the video.  All panels: error ribbons denote bootstrap-estimated standard error of the mean.}
  \label{fig:list-learning}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The video-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of 1 minus the average correlation between each recall event and all other recall events that do not display a box in the same row.  \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{D.} The correlation between precision and the number of video events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of video events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}
% Need to reconcile the description of distinctiveness with how we computed it... - Andy

Statistical models of memory studies often treat recall success as binary (in other words, an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, memory for naturalistic experiences is much more nuanced.  For example, certain aspects of an experience might be correctly remembered at varying levels of detail, or distorted, or forgotten entirely.  Further, each remembering is itself a richly structured phenomenon.  Our framework produces a content-based model of individual video and recall events by projecting the dynamic content of the video and participants' recalls into a shared topic space.  This allows for direct, quantitative comparisons between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  Precision is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in memory.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any video event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recalled event reflects the most similar presented event moreso than it does other presented events.  To compute a recall event's distinctiveness, we first identify the video event to which its topic vector is most strongly correlated.  We then define distinctiveness as one minus the average correlation between the given recall event and all \textit{other} video events.  In addition to individual events, one may also use these metrics to describe each participant's overall performance by averaging across a participant's event-wise precision or distinctiveness scores.  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, higher precision scores are positively correlated with both hand-annotated memory performance (as collected by Chen et al., 2017; Pearson's $r(15) = 0.60, p = 0.010$) and the number of video events successfully remembered, as determined by our model (Pearson's $r(15) = 0.90, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that participants' distinctiveness scores were positively
correlated with our model's estimated number of recall events (Pearson's $r(15) = 0.55, p = 0.021$).  However, we found no evidence that distinctiveness scores were correlated with hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$).  We elaborate on this potential discrepency in the \textit{Discussion} section.

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision metric reflects completeness of recall.} \textbf{A.} Recall precision by video event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single video event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  Video events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details" video annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example video event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} A subset of the sentences comprising the most precise (P17) and least precise (P6) participants' recalls of video event 22.  Descriptions of specific actions or features reflecting those highlighted in panel B are highlighted in the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/distinctiveness_detail}
  \caption{\small \textbf{Distinctiveness metric reflects specificity of recall.} \textbf{A.} Recall distinctiveness by video event.  Kernel density estimates for each video event's distribution of recall distinctiveness scores, analogous to Fig.~\ref{fig:precision-detail}A.  \textbf{B.} The sets of ``Narrative Details" video annotations (generated by \citealp{ChenEtal17}) for scenes comprising video events described by the example participants in panel C.  Each event's text is highlighted in a different color.  \textbf{C.} The sentences comprising the most distinctive (P13) and least distinctive (P11) participants' recalls of video event 19.  Sections of recall describing each each video event in panel B are highlighted with the corresponding color.}
  \label{fig:distinctiveness-detail}
\end{figure}

>>>>

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 scenes, and treated each window as a single ``document" for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used hand-annotated descriptions of each manually identified scene from the episode to fit a topic model.  Three example video frames (first row) and their associated descriptions (second row) are displayed.  The third row shows an example participant's later recalls of the same three scenes.  We used the topic model (fit to the episode annotations) to estimate topic vectors for each moment of the episode and each sentence of participants' recalls.  Example topic vectors are displayed in the bottom row (blue: episode annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively), along with the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% add an x-label under the "

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number (typically one or two) of topics tended to be ``active" in any given timepoint (see Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events" into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 episode timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 episode events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the episode.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recall events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of episode events (Panel C) and recall events (from the example participant; Panel F).  For similar plots for all participants, see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of episode events and recall events (across all 17 participants).  To create the figure, each recalled event was assigned to the episode event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}
% I'm all for citing software packages, but where what's the distinction between packages we do cite (seaborn, brainiak, sklearn, hypertools, quail, umap, etc.) and don't cite (wordcloud, nilearn, fastdtw, etc.)?
% - Paxton
% I also flagged the seaborn reference for removal here. Def not critical to understanding the figure - Andy

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall, we treated each overlapping  window of (up to 10) sentences from their transcript as a ``document," and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite different participants using widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This is a substantial benefit of projecting the episode and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall trajectory (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the episode correlation matrix, we leveraged an HMM-based optimization procedure (see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  Whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results indicate that both the structure of the original episode and participants' recalls of the episode exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events via HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the episode, and the events they later recalled.  One approach to linking the experienced (episode) and recalled events is to label each recalled event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented" events from the original episode, and a (potentially differently ordered) sequence of ``recalled" events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observed patterns comparable to classic effects from list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). In contrast, we did not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather greater memory for specific events distributed approximately evenly throughout the episode.

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.~Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events were clustered heavily in participants' recall event sequences (mean clustering score: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.
% NOTE:
% Semantic and temporal clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.
% - Paxton

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the episode. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the episode.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the episode.  All panels: error ribbons denote bootstrap-estimated standard error of the mean.}
  \label{fig:list-learning}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of 1 minus the average correlation between each recall event and all other recall events that do not display a box in the same row.  \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{D.} The correlation between precision and the number of episode events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of episode events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}
% Need to reconcile the description of distinctiveness with how we computed it... - Andy

Statistical models of memory studies often treat recall success as binary (in other words, an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, memory for naturalistic experiences is much more nuanced.  For example, certain aspects of an experience might be correctly remembered at varying levels of detail, or distorted, or forgotten entirely.  Further, each remembering is itself a richly structured phenomenon.  Our framework produces a content-based model of individual episode and recall events by projecting the dynamic content of the episode and participants' recalls into a shared topic space.  This allows for direct, quantitative comparisons between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  Precision is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in memory.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any episode event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recalled event reflects the most similar presented event moreso than it does other presented events.  To compute a recall event's distinctiveness, we first identify the episode event to which its topic vector is most strongly correlated.  We then define distinctiveness as one minus the average correlation between the given recall event and all \textit{other} episode events.  In addition to individual events, one may also use these metrics to describe each participant's overall performance by averaging across a participant's event-wise precision or distinctiveness scores.  Participants whose recall events are more veridical descriptions of what happened in the episode event will presumably have higher precision scores. We find that, across participants, higher precision scores are positively correlated with both hand-annotated memory performance (as collected by Chen et al., 2017; Pearson's $r(15) = 0.60, p = 0.010$) and the number of episode events successfully remembered, as determined by our model (Pearson's $r(15) = 0.90, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that participants' distinctiveness scores were positively
correlated with our model's estimated number of recall events (Pearson's $r(15) = 0.55, p = 0.021$).  However, we found no evidence that distinctiveness scores were correlated with hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$).  We elaborate on this potential discrepency in the \textit{Discussion} section.

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision metric reflects completeness of recall.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  episode events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details" episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} A subset of the sentences comprising the most precise (P17) and least precise (P6) participants' recalls of episode event 22.  Descriptions of specific actions or features reflecting those highlighted in panel B are highlighted in the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/distinctiveness_detail}
  \caption{\small \textbf{Distinctiveness metric reflects specificity of recall.} \textbf{A.} Recall distinctiveness by episode event.  Kernel density estimates for each episode event's distribution of recall distinctiveness scores, analogous to Fig.~\ref{fig:precision-detail}A.  \textbf{B.} The sets of ``Narrative Details" episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in panel C.  Each event's text is highlighted in a different color.  \textbf{C.} The sentences comprising the most distinctive (P13) and least distinctive (P11) participants' recalls of episode event 19.  Sections of recall describing each each episode event in panel B are highlighted with the corresponding color.}
  \label{fig:distinctiveness-detail}
\end{figure}
